Bicycle Balance Assistance Reduces Probability of Falling at Low Speeds When Subjected to Mechanical Perturbations

### 2024-04-12 main.tex
Editing of most of the sections.

\section{Introduction}
%
Single actor bicycle crashes are associated with a large number of unreported
and reported injuries and even sometimes death. At low speeds (less than 15
km/h or so) bicycles are no longer self stable and become more difficult to
control for the rider. Low speed crashes may be reduced if the bicycle was self
stable. Since at least the 1980s it known that applying a steering torque
proportional to the roll rate can stabilize a single track vehicle at very low
speeds. If the robotic control of steering can stabilize a bicycle, it may
reduce the control required by the rider to successfully manage balancing
tasks. We have developed such a balance assisting bicycle and desire to
evaluate whether it helps the rider in situations they are likely to fall.

\section{Methods}
%
\subsection{Bicycle}
%
We modified a standard electric bicycle (Gazelle XXX, Dieren, The Netherlands)
with a custom motor in the head tube capable of applying up to
7~\si{\newton\meter} of torque between the head tube and steer tube. A custom
motor controller converts the commanded torque to applied torque. We measure
the speed of the rear wheel with an encoder (Magura XXX) and measure the body
fixed roll rate of the bicycle with a MEMs rate gyroscope. The balance assist
control algorithm is implemented on microprocessor (Teensy) and data from all
sensors is logged with a CAN bus.

\subsection{Balance Assist Control}
We use a forward speed \(v\) gain scheduled proportional roll rate positive
feedback controller to stabilize the bicycle. The commanded steer torque
\(T_\delta\) from the steer motor follows the control law
\begin{align}
  T_\delta = k(4.7\si{\meter\per\second} - v)\dot{\phi}
\end{align}
where 4.7~\si{\meter\per\second} is the weave critical speed predicted from the
open loop bicycle rigid rider dynamics. We use \(k=8\) as the low gain value
and \(k=10\) as a high gain value. Scaling the proportional feedback gain
linearly with respect to speed stabilizes the normally unstable weave mode of
the bicycle from speeds of X~\si{\kilo\meter\per\hour} to
X~\si{\kilo\meter\per\hour} as shown in Figure~\ref{fig:eigenvalues}.

@@

\subsection{Perturbations}
%
We apply external longitudinal forces to the ends of each handlebar using an
adapted Bump'Em system~\cite{Tan2020} which is designed with four motors
working in tandem. The four motors are programmed to apply a light force at all
times to keep the ropes taught and to track a commanded force profile using a
PID controller running on a microprocessor (Arduino XXX, Italy). We measure the
force applied by each motor at the handlebar via four inline load cells with a
max load of XX~\si{\newton}. The commanded force profiles are designed to apply
an external pulsive torque to the front assembly (handlebars, forks, wheel) at
magnitudes varying from X to X. The four motors are arranged at the four
corners of a 1~\si{\meter} wide treadmill (make, Company, City) that can reach
speeds of 18~\si{\kilo\meter\per\hour}. The general design is described in
detail in the Van De Velde's MSc thesis~\cite{vandeVelde2022}. Our
modifications to that described in the thesis include simplifying the
controller with a inexpensive microcontroller and the use of a simpler safety
harness.

@@

\subsection{Two Experiments}
%
The 10~\si{\kilo\meter\per\hour} experiments occurred X weeks before the
6~\si{\kilo\meter\per\hour}.

\subsection{Protocol}
%
Subjects wore a helmet and fall safety harness attached to the ceiling. We
allowed them practice riding on the treadmill until they indicated they were
comfortable enough to apply perturbations. For most this was less than a
10~\si{\minute} warm up. We then ask the rider to ride for 90~\si{\second}
attempting to maintain the location of their front wheel on the center line of
the treadmill. We define a "fall" on the treadmill by two criteria: the rider
removes their foot from the pedal and places it on the ground and the bicycle
wheel exceeds the width of the treadmill. We then applied perturbations in
random directions (clockwise or counter clockwise) increasing the magnitude by
X~\si{\newton\meter} each time. We log the magnitude that causes the first fall
to characterize that subject's nominal fall threshold. Following this we apply
20 perturbations of random magnitudes to the cyclist while they ride at a
constant speed and record which perturbations cause a fall. We let the cyclist
rest and then perform another 20 perturbations. We randomize whether the
balance assist system is on or off during the first or second set of
perturbations.

>>>>

\section{Introduction}
%
Single actor bicycle crashes are associated with a suprisingly large number of
reported injuries~\citep{Wegman2024}. At low speeds, from start up to typical
cruising speeds, bicycles themselves are not self-stable and are challenging to
control for the rider. Low speed crashes may be reduced if the bicycle was
self-stable at these speeds. Bicycles can be mechanically modified to lower the
speeds at which it is self-stable~\citep{Astrom2005} but since at least the
1980s it is known that applying a steering torque proportional to the roll rate
can stabilize a single track vehicle at very low speeds. If the robotic control
of steering can stabilize a bicycle, it may reduce the control required by the
rider to successfully manage balancing tasks. We have developed such a balance
assisting bicycle and hypothesize that it helps the rider in situations they
are likely to fall.

Automatic roll stabilization of single track vehicles began in earnest after
predictive motorcycle models were developed and refined throughout the 1970s.
\Citet{vanZytveld1975} seems to be the first
\todo{Check if a Japanese researcher was first}
to attempt to robotically stabilize a small motorbike with a controlled
inverted pendulum that mimicked rider lean, but he was not successful in
demonstrating what his control model predicted. In his model, he recognized
that feedback of the vehicle roll angle and angular rate was essential to
stabilize the vehicle. It was not until the mid 1980s that someone successfully
demonstrated an automatically balanced motorcycle~\citep{Ruijs1986a} and they
did so with a steering motor. Ruijs and Pacejka clearly showed that steer
torque driven by roll angle feedback stabilizes the capsize mode, roll angular
rate feedback stabilizes the weave mode, and steer angular rate feedback
stabilizes the wobble mode. They also show how the gains need to change with
respect to vehicle speed for favorable control at all speeds. This roll motion
feedback is the simplest controller that can stabilize a single track vehicle
above a minimum speed when not concerned with wobble instabilities,
which are not prominent in bicycles. Ruijs and Pacejka's work was not
particularly concerned with low speed stability and their vehicle was fully
automatic, no human rider was involved.  Many more automatically balanced
single track vehicles have been demonstrated over the last 40 years, but none
have demonstrated that increasing low speed stability can assist in balancing
and possibly reduce single actor falls. Most of these robotic bicycles and
motorcycles designers did not intend for a human rider to also control the
stabilized vehicle. Nevertheless an automatically stabilized bicycle can be
controlled by a human rider if the motor controlled steer torque and the rider
applied steer torque act on the steer in parallel. The effect of this automatic
control gives the ability to change the dynamics, up to some limits, of the
human controlled plant.

The linear Carvallo-Whipple bicycle model~\cite{Carvallo1899,Whipple1899} is
the simplest bicycle model that exhibits non-minimum phase behavior and
self-stablity. Interestingly, Whipple showed that roll motion feedback
stabilizes the system back in 1899 and the model is ideal for showing the
effect of a roll motion feedback driven steer motor on the dynamics. The linear
model can be described by the fourth order state space equations:

\begin{align}
  \dot{\vec{x}} = \mathbf{A} \vec{x} +
  \mathbf{B}
  \vec{u}
  \textrm{ where }
  \vec{x} = \begin{bmatrix} \phi \\ \dot{\phi} \\ \delta \\ \dot{\delta} \end{bmatrix}
  \textrm{ and }
  \vec{u} = \begin{bmatrix} T_{\phi} \\ T_{\delta} \end{bmatrix}
\end{align}

The states are the roll angle \(\phi\) and steer angle \(\delta\) along with
their derivatives and the inputs are roll torque \(T_\phi\) and steer torque
\(T_\delta\). The state \(\mathbf{A}\) and input \(\mathbf{B}\) matrices are
functions of the equilbirum forward speed \(v\) and  are otherwise populated
with expressions that are functions of the geometric and inertial parameters of
the nonholonomic mulitbody system made up of four rigid bodies: two wheels,
front frame, and rear frame.

If the steer torque is the sum of the (h)uman applied torque and the (m)otor
applied torque \(T_\delta = T_\delta^\textrm{h} + T_\delta^\textrm{m}\),
\(\mathbf{B} = \begin{bmatrix} \mathbf{B}_\phi \quad \mathbf{B}_\delta
\end{bmatrix}\), and \(T_\delta^\textrm{m} = -k_{\dot{\phi}} \dot{\phi}\) then
the human controlled plant takes the form:

\begin{align}
  \dot{\vec{x}} = \left(\mathbf{A} -
  \mathbf{B}_\delta
  \left[0 \quad k_{\dot{\phi}} \quad 0 \quad 0\right]
\right)
  \vec{x} + \mathbf{B} \begin{bmatrix} T_{\phi} \\ T_\delta^\textrm{h} \end{bmatrix}
\end{align}

The state matrix \(\mathbf{A}\) and input matrix \(\mathbf{B}\) are both
functions of the equilibrium speed \(v\) and \(k_{\dot{\phi}}\) can be selected
such that the eigenvalues of \(\left(\mathbf{A} - \mathbf{B}_\delta\left[0 \quad k_{\dot{\phi}}
\quad 0 \quad 0\right] \right)\) have negative real parts for
\(v_{weave} < v < v_{capsize}\). With gain scheduling with respect to \(v\),
the speed range where the bicycle is stable can be maximized within
any physical acutator magnitude and bandwidth limits. \citet{Schwab2008}
elaborates on some of the possibilities in scheduling the gains for such a
controller.

\todo[inline]{I'm not consistent with the arrow being a vector for $B_\delta$}

In this paper, we test whether a gain scheduled steer motor controlled bicycle
that is stable at a low speed range when riderless assists in balancing by
investigating whether it is beneficial in preventing the rider from falling.
To do this, we apply varying magnitude mechanical perturbations to the
handlebars while the rider is balancing the bicycle on a treadmill and then
assess the rider's probability of falling with the balance assist steer motor
system on and off.

\section{Methods}
%
\subsection{Bicycle}
%
We modified a standard electric bicycle (Gazelle XXX, Dieren, The Netherlands)
with a custom motor in the steering assembly capable of applying up to
7~\si{\newton\meter} of torque between the head tube and steer tube. A custom
motor controller converts the commanded torque to applied torque. We measure
the speed of the rear wheel with an encoder (Magura XXX) and measure the body
fixed roll rate of the bicycle with a MEMs rate gyroscope (Company, Location).
The balance assist control algorithm is implemented on a microprocessor
(Teensy, Company, Location) and data from all sensors is logged with a CAN bus
(Company, Location) at approximately X Hz.

\subsection{Balance Assist Control}
We use a forward speed \(v\) gain scheduled proportional roll rate feedback
controller to stabilize the bicycle. The commanded steer torque
\(T^\textrm{m}_\delta\) from the steer motor follows the control law
\begin{align}
  T^\textrm{m}_\delta = k(v_{weave} - v)\dot{\phi}
\end{align}
where 4.7~\si{\meter\per\second} is the weave critical speed predicted from the
open loop bicycle rigid-rider dynamics. We use \(k=8\) and \(k=10\) as gain
values during the experiments. Scaling the proportional feedback gain linearly
with respect to speed stabilizes the normally unstable weave mode of the
bicycle from speeds of X~\si{\kilo\meter\per\hour} to
X~\si{\kilo\meter\per\hour} as shown in Figure~\ref{fig:eigenvalues}.

@@

\subsection{Perturbations}
%
We apply external longitudinal forces to the ends of each handlebar using an
adapted Bump'Em system~\cite{Tan2020} which is designed with four motors
working in tandem. The four motors are programmed to apply a light force at all
times to keep the ropes taught and to track a commanded force profile using a
PID controller running on a microprocessor (Arduino XXX, Italy). We measure the
force applied by each motor at the handlebar via four inline load cells, each
with a max load of XX~\si{\newton}. The commanded force profiles are designed
to apply an external pulsive torque to the front assembly (handlebars, forks,
wheel) at magnitudes varying from X to X. The four motors are arranged at the
four corners of a 1~\si{\meter} wide treadmill (make, Company, City) that can
reach speeds of 18~\si{\kilo\meter\per\hour}. The general design is described
in detail in the Van De Velde's MSc thesis~\cite{vandeVelde2022}. Our
modifications relative to that described in the thesis include simplifying the
controller with a inexpensive microcontroller and the use of a simpler safety
harness.

@@

\subsection{Protocol}
%
The subjects were divided into two groups, where the first group performed the
protocol at a belt speed of 10~\si{\kph} with the controller gain as $k=8$ and
the second group performed the protocol at a belt speed of 6~\si{\kph}. The
10~\si{\kph} group's experiments occured X weeks before the
6~\si{\kilo\meter\per\hour} group.

Subjects wore a helmet and fall safety harness attached to the ceiling. We
allowed them practice riding on the treadmill until they indicated they were
comfortable enough to have perturbations applied. For most, this was less than a
10~\si{\minute} warm up. We then asked the rider to ride for 90~\si{\second},
attempting to maintain the location of their front wheel on the center line of
the treadmill. We define a "fall" on the treadmill by two criteria: the rider
removes their foot from the pedal and places it on the ground or the bicycle
wheel exceeds the width of the treadmill. We then applied perturbations in
random directions (clockwise or counter clockwise) increasing the magnitude by
X~\si{\newton\meter} each time. We log the magnitude that causes the first fall
to characterize that subject's nominal fall threshold. Following this we apply
20 perturbations of random magnitudes to the cyclist while they ride at a
constant speed and record which perturbations cause a fall. We let the cyclist
rest and then perform another 20 perturbations. We randomize whether the
balance assist system is on or off during the first or second set of
perturbations.
